---
title: "Dataset manipulation pt 2"
author: "Amanda Mae Woodward"
date: "2024-10-15"
output: html_document
---

We're going to use this part of class to review how to do the learning outcomes from the dataset manipulation notes. 

### load data
1. Open the palmerpenguins library
```{r}
library(palmerpenguins)
#remotes::install_github("allisonhorst/palmerpenguins")
```

2. Open the penguins dataset
```{r}
data(penguins)
```

3. Explore the raw dataset. What can you find out about it? 
look in the help file; we agree that penguins dataset looks better 

4. Are the data in long or wide format? 

```{r}
summary(as.factor(penguins_raw$`Sample Number`))
summary(as.factor(penguins_raw$Species))
summary(as.factor(penguins_raw$`Individual ID`)) #this is what we'll use 
```


5. let's practice going from one format to the other. 
long to wide: 

using flipper length as the variable
```{r}
library(tidyverse)

tinyPenguin<- penguins_raw %>% select("Individual ID","Sample Number", "Flipper Length (mm)")
```
Note: individual id is not nested within sample number. we can still run this code, but it's gonna be messy. (making data messier is a thing we can do- just not normally the goal)
```{r}
tinyPenguinWide <- pivot_wider(tinyPenguin, id_cols = "Individual ID",names_from= "Sample Number", values_from=  "Flipper Length (mm)") 

tinyPenguinWide <- tinyPenguin %>% pivot_wider(id_cols = "Individual ID",names_from= "Sample Number", values_from=  "Flipper Length (mm)") 

tinyPenguin2<- spread(penguins_raw, key= "Sample Number", value = "Flipper Length (mm)")
```


```{r}
penguinLongData<- penguins_raw %>% pivot_longer(cols = c("Culmen Length (mm)", "Culmen Depth (mm)","Flipper Length (mm)","Body Mass (g)"), names_to = "bodyMeasures", values_to= "measurements")
```

```{r}
penguinWide<- penguinLongData %>% pivot_wider(id_cols = c("Individual ID", "Sample Number"), names_from = "bodyMeasures", values_from="measurements")
```


6. create two datasets. One should have the individual id, region, and island. The other should have individual id, species, flipper length, culmen length, culmen depth, and body mass. 

```{r}
penguinA<- penguins_raw %>% select("Individual ID", "Region", "Island")
penguinB<- penguins_raw %>% select("Individual ID", "Culmen Length (mm)", "Culmen Depth (mm)","Flipper Length (mm)","Body Mass (g)")
```

7. Join the two datasets you made together.
```{r}
a<- left_join(penguinA, penguinB)
```

8. Separate the species column into the common name and the scientific name 
```{r}
penguins_raw$Species<- as.factor(penguins_raw$Species)
levels(penguins_raw$Species)<- c("Adelie Penguin, Pygoscelis adeliae", "Chinstrap penguin, Pygoscelis antarctica","Gentoo penguin, Pygoscelis papua")
b<-separate(penguins_raw,col = 'Species',into= c("commonName", "scientificName"), sep = ",")
```

9. change the levels of clutch completion to indicate whether the penguin has two eggs or does not have two eggs

```{r}

```

10. create a column that has body mass in kilograms.

```{r}

```

11. copy island to a new column. Recode this variable so that each island is renamed to be just the first initial

```{r}

```

